---
title: "Task 2"
author: "Grace Bianchi"
date: "2023-03-01"
output: 
  html_document:
    code_folding: hide
  
---

```{r setup, include=TRUE, warning = FALSE, message = FALSE, echo = TRUE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(here)
library(tidyverse)
library(ggdendro)

```

### Introduction

This code explores Santa Barbara stream data using hierarchical clustering by complete linkage. A dendrogram is created to shower the clustering for the stream chemistry by site. This data includes stream chemistry in the Santa Barbara Coastal drainage area since 2000. Samples from the watersheds includes measurements of dissolved nitrogen (nitrate, ammonium, total dissolved nitrogen); soluble reactive phosphorus (SRP); particulate organic carbon, nitrogen and phosphorus; total suspended sediments; and conductivity.


**Citation:** Santa Barbara Coastal LTER and J. Melack. 2019. SBC LTER: Land: Stream chemistry in the Santa Barbara Coastal drainage area, ongoing since 2000 ver 16. Environmental Data Initiative. https://doi.org/10.6073/pasta/67a558a24ceed9a0a5bf5e46ab841174. 


```{r data wrangling}
stream_data <- read_csv(here("sbc_lter_registered_stream_chemistry.csv")) %>%
  janitor::clean_names() %>% 
  mutate(across(where(is.numeric), ~na_if(., -999.0)))  # convert -999 to NA

stream_nas_by_variable <- stream_data %>% 
  summarize(across(everything(), ~sum(is.na(.))))
```

After exploring the stream data, it appears the total particulate carbon (TPC), total particulate nitrogen, and total particulate phosphorus had the greatest number of missing values, 15,521, 16,521, and 17,046, respectively. The absences of these values were mostly consistent across each site and more than 50% of the values in each variable were NA, thus these three variables were removed from further analysis. In addition, the total suspended solids (TSS) data was removed, as more 50% of the values were NA. 

```{r}
# calculate mean concentrations for each site
stream_avg <- stream_data %>% 
  group_by(site_code) %>% 
  summarize(mean_nh4 = mean(nh4_u_m, na.rm = TRUE),
            mean_no3 = mean(no3_u_m, na.rm = TRUE),
            mean_po4 = mean(po4_u_m, na.rm = TRUE),
            mean_tdn = mean(tdn_u_m, na.rm = TRUE),
            mean_tdp = mean(tdp_u_m, na.rm = TRUE),
            mean_spec = mean(spec_cond_u_spercm, na.rm = TRUE)) 
```

### Hierarchicial Clustering

The euclidean distance is calculated before performing the complete linkage agglomerative hierarchical clustering.
This is a bottom-up approach in which all observations start as their own cluster and then are grouped based on similarity.

```{r}
## scale data
streams_scaled <- stream_avg %>% 
  select(where(is.numeric)) %>% 
  scale()

rownames(streams_scaled) <- stream_avg$site_code 

streams_dist <- dist(streams_scaled, method = 'euclidean') # euclidean distance

# hierarchical clustering complete linkage
streams_hc_complete <- hclust(streams_dist, method = 'complete')

ggdendrogram(streams_hc_complete) +
  theme_minimal() +
  labs(x = 'Stream Site', y = 'Height')
```

**Figure 1.** Hierarchical clustering by complete linkage of Santa Barbara stream chemistry. The clusters/grouping indicate how closely each stream site is to one another.The height represents the distance between clusters.

**Summary**

- MC00 & SP02 and HO00 & RS02 are the most similar to each other
- There are two distinct clusters at a height around 3
- DV01 and BC02 are the most different from the other stream sites
