---
title: "task 1"
author: "Grace Bianchi"
date: "2023-03-01"
output: html_document
---


```{r setup, include=TRUE, warning = FALSE, message = FALSE, echo = TRUE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(here)
library(tidyverse)
library(broom)

# For Part 1 (PCA)
library(ggfortify) # For PCA biplot
library(patchwork)
```

Principal Component Analysis

Intro Summary (3-4 sentences)

What are we exploring: 

The dataset used in this code includes environmental variables for 243 countries, in which most of the data was "derived by calculating the mean for each country at a reduction scale of about 10km.” This data was compiled and provided by @zander_venter on Kaggle, where all data was acquired through Google Earth Engine (https://earthengine.google.com/), where publicly available remote sensing datasets have been uploaded.



**Citation:**  



```{r wrangle data}
env_vars_data <- read_csv(here("world_env_vars.csv")) %>% 
  janitor::clean_names() %>% 
  # summarize(across(everything(), ~sum(is.na(.))))  number of NAs in each col
  select(!wind) %>%  # removing quarterly data since there are many 
  drop_na()

env_vars_pca <- env_vars_data %>% 
  select(where(is.numeric)) %>% # select only numeric columns for PCA 
  scale() %>% 
  prcomp() 

pca_sum <- summary(env_vars_pca)  


```


### Principle Component Analysis

Principle component 1 captures 41% of variance, followed by principle component 2 which captures 24% variance, and 95% of the variance is captured by the first 10 principle components.

```{r}
# See the loadings (weighting for each principal component)
env_vars_pca$rotation
```

#### PCA BiPlot

```{r}

## PCA Bi Plot
autoplot(env_vars_pca,
         data = env_vars_data,
         loadings = TRUE,
         color = "country",
         loadings.label = TRUE,
         loadings.color= "black",
         loadings.label.vjust = -0.5) +
  theme_minimal()
```

**Figure Caption.** 

### Summary

Major Take aways

