---
title: "Task 2"
author: "Grace Bianchi"
date: "2023-03-01"
output: 
  html_document:
    code_folding: hide
  
---

```{r setup, include=TRUE, warning = FALSE, message = FALSE, echo = TRUE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(here)
library(tidyverse)
library(kableExtra)

```

### Introduction


A useful descriptive introductory summary (3 - 4 sentences) that helps the audience understand the data (include a citation as necessary) and what you’ll be exploring

This data includes stream chemistry in the Santa Barbara Coastal drainage area since 2000. Samples from the watersheds includes measurements of dissolved nitrogen (nitrate, ammonium, total dissolved nitrogen); soluble reactive phosphorus (SRP); particulate organic carbon, nitrogen and phosphorus; total suspended sediments; and conductivity.


**Citation:** Santa Barbara Coastal LTER and J. Melack. 2019. SBC LTER: Land: Stream chemistry in the Santa Barbara Coastal drainage area, ongoing since 2000 ver 16. Environmental Data Initiative. https://doi.org/10.6073/pasta/67a558a24ceed9a0a5bf5e46ab841174. 


```{r data wrangling}
stream_data <- read_csv(here("sbc_lter_registered_stream_chemistry.csv")) %>% 
  janitor::clean_names() %>% 
  mutate(across(where(is.numeric), ~na_if(., -999.0)))  # convert -999 to NA


stream_nas_by_variable <- stream_data %>% 
  summarize(across(everything(), ~sum(is.na(.))))

stream_nas_by_site <- stream_data %>% 
  group_by(site_code) %>% 
  summarize(across(everything(), ~sum(is.na(.))))

```


After exploring the stream data, it appears the total particulate carbon (TPC), total particulate nitrogen, and total particulate phosphorus had the greatest number of missing values, 15,521, 16,521, and 17,046, respectively. The absences of these values were mostly consistent across each site and more than 50% of the values in each variable were NA, thus these three variables were removed from further analysis. In addition, the total suspended solids (TSS) data was removed, as more 50% of the values were NAs. Specific conductivity was removed because it was not useful for this analysis.

```{r}

stream_tidy <- stream_data %>% 
  select(!c("tpp_u_m", "tpc_u_m", "tpn_u_m", "tss_mgper_liter", "spec_cond_u_spercm")) %>% 
  pivot_longer(cols = where(is.numeric),
               names_to = "variable",
               values_to = "value")

```



 
 Exploratory Visualizations
 
```{r}
ggplot(data = stream_tidy, aes(x = timestamp_local, y = value)) +
  geom_line(aes(color = variable)) +
  facet_wrap(~site_code)

  
ggplot(data = stream_tidy) +
  geom_boxplot(aes(x = variable, y = value))
```


To perform hierarchical clustering by site, you’ll want to make a data frame that has a single summary row per site (e.g. based on means from all observations at that site), then calculate the Euclidean distance before performing complete linkage agglomerative hierarchical clustering. 

```{r}

stream_summary <- stream_tidy %>% 
  group_by(site_code, variable) %>% 
  summarize(mean_nh4 = mean(nh4_u_m, na.rm = TRUE),
            mean_no3 = mean(no3_u_m, na.rm = TRUE),
            mean_po4 = mean(po4_u_m, na.rm = TRUE),
            mean_tdn = mean(tdn_u_m, na.rm = TRUE),
            mean_tdp = mean(tdp_u_m, na.rm = TRUE))
  
  
```


```{r}
# single summary per row per site 
stream_sum <- stream_data %>% 
  group_by(site_code) %>% 
  summarize(mean_nh4 = mean(nh4_u_m, na.rm = TRUE),
            mean_no3 = mean(no3_u_m, na.rm = TRUE),
            mean_po4 = mean(po4_u_m, na.rm = TRUE),
            mean_tdn = mean(tdn_u_m, na.rm = TRUE),
            mean_tdp = mean(tdp_u_m, na.rm = TRUE))

kable(stream_sum)
```







```{r}
streams_dist <- dist(streams_scale, method = "euclidean")

streams_hc_complete <- hclust(streams_dist, method = "complete")

plot(streams_hc_complete, ces = 0.6, hang = -1)
```


All of your organized and well-annotated code (with warnings/messages suppressed) you wrote to wrangle data then create the dendrogram, and to create a professional looking dendrogram that appears (with a caption) in your knitted HTML.  


A brief summary (nicely formatted and professionally written bulletpoints are fine) highlighting some major takeaways from your hierarchical clustering analysis that can be gleaned from the dendrogram(s).


